---
title: "Probability with Discrete Random Variables"
author: "Ben Goodrich"
format:
  revealjs:
    embed-resources: true
    self-contained-math: true
editor: visual
engine: knitr
execute: 
  echo: true
preload-iframes: false
---

## Jobs, Jobs, Jobs

According to one of Gelman's blog posts [today](https://statmodeling.stat.columbia.edu/2025/07/08/52215/), the Los Angeles Dodgers are interested in hiring someone with

> demonstrated strength in Bayesian hierarchical modeling & probabilistic forecasting \[meaning those who\] have written probabilistic models in NumPyro, PyMC, or Stan with custom likelihoods and priors

. . .

Why?

## Goals for Today

-   Start to learn probability with some degree of rigor

-   Compare and contrast ancient and modern computation

    -   Ancient Bayesian inference could be done with a pen but was feasible only for a small subset of problems

    -   Modern Bayesian inference (1990+) is mostly done with computer simulations, which is possible with R for a large subset of problems but can be slow with Big Data

    -   When Ancient Bayesian inference is feasible (e.g., today), it yields essentially the same answer as Modern

. . .

Make pairs with 1 from statistics and 1+ who knows dplyr

## Random Variables (R.V.)

-   A function is a rule that *uniquely* maps each element of an input set to some element of an output set, e.g. $e^x$ maps real numbers $\left(\mathbb{R}\right)$ to non-negative real numbers $\left(\mathbb{R_+}\right)$
-   A random variable is a *function* from the sample space, $\Omega$, to some subset of $\mathbb{R}$ with a probability-based rule
-   If $\Omega$ is discrete with a finite number of elements, then we can simply enumerate an equivalent number of probabilities

::: panel-tabset
### R

```{r}
x <- sample(1:6, size = 1, prob = rep(1 / 6, times = 6)) # do this
```

### Python

```{python}
from sympy.stats import *
X = Die('X', 6) # RV
x = sample(X)   # a number
```
:::

## Bowling Basics

Each "frame" in bowling starts with $n = 10$ pins. You get up to two rolls per frame to knock down as many pins as you can. {{< video https://www.youtube.com/embed/HeiNrSllyzA width="800" height="400" >}}

## Approaching Bowling Probabilistically

::: incremental
-   What is $\Omega$ for the number of pins knocked down on roll $1$?
-   If $b^p = y$, then $\log_b\left(y\right) = p$. Let the probability of knocking down $x$ out of $n$ pins be given by a form of [Benford's Law](https://en.wikipedia.org/wiki/Benford%27s_law): $\Pr\left(x \mid n\right) = \log_{n + 2}\left(1 + \frac{1}{n + 1 - x}\right)$, presuming $0 \leq x \leq n$.
:::

. . .

```{r, Pr}
# probability of knocking down x out of n pins
Pr <- function(x, n = 10) ifelse(x > n, 0, log(1 + 1 / (n + 1 - x), n + 2))
Omega <- 0:10 # 0, 1, ..., 10
names(Omega) <- as.character(Omega)
source("bowling.R") # does the above (or look at bowling.py)
x_1 <- sample(Omega, size = 1, prob = Pr(Omega)) # do this
```

. . .

```{r}
round(c(Pr(Omega), total = sum(Pr(Omega))), digits = 4)
```

## Second Roll in a Frame of Bowling

-   How would you draw $x_2$, which is the realization of your second roll in the first frame of bowling?

. . .

```{r}
x_2 <- sample(Omega, size = 1, prob = Pr(Omega, n = 10 - x_1)) # legal?
```

. . .

-   If $x_1 > 0$, some elements of $\Omega$ have zero probability in the second roll, which is enforced by the `ifelse()` in `Pr()`
-   $\Pr\left(x \mid n = 10 - x_1\right)$ is a *conditional* probability because it depends on the realization of $x_1$ via $n = 10 - x_1$
-   [Joe Blitzstein](https://youtu.be/dzFf3r1yph8): "Conditioning is the soul of statistics"

. . .

-   Pairs exercise: Simulate $R = 10$ frames of bowling (each with two rolls), storing the number of pins knocked down

## Simulating $R$ Frames of Bowling

```{r}
#| message: false
#| code-line-numbers: 1-2|3-4|5-8
R <- 10^7  # practically infinite (to almost match exact calculations)
frames <-  # tibble with the results of R frames of bowling
  tibble(X_1 = sample(Omega, size = R, replace = TRUE, 
                      prob = Pr(Omega))) |> # all R first rolls
  group_by(X_1) |> # then all second rolls, one group at a time
  mutate(X_2 = sample(Omega, size = n(), replace = TRUE, 
                      prob = Pr(Omega, n = 10 - first(X_1)))) |>
  ungroup() # first(X_1) is needed so that n is a scalar in ^^^
```

. . .

```{r}
print(frames, n = 10)
```

## Probability Via dplyr / siuba Syntax

`|>` pipes a previously created `tibble` to the next function

| Concept             | R Syntax \[after callling **library(dplyr)**\] |
|---------------------|------------------------------------------------|
| $X$                 | `tibble(X = sample(…))` or `|> mutate()`       |
| $\mid X = x$        | `|> filter(X == x) |> ...`                     |
| $\mid X$            | `|> group_by(X) |> ... |> ungroup()`           |
| $\Pr\left(x\right)$ | `|> summarize(prob = mean(X == x))`            |

Also, `count()` is essentially equivalent to `group_by(X) |> summarize(times = n())` where the `n()` function yields the number of observations for each group in the `tibble`

## Checking the Simulations

```{r}
frames |> # vvv is like group_by(frames, X_1) |> summarize(times = n())
  count(X_1, name = "times") |>  # a tibble with 11 rows
  mutate(proportion = times / R, # almost exact but not quite
         probability = Pr(X_1))  # same concept as proportion but exact
```

## Joint (here bivariate) Probabilities

-   How would you compute the probability that $X_1 = 8$ and $X_2 = 2$ in a frame of bowling from the simulated `frames`?

. . .

```{r}
summarize(frames, joint_prob = mean(X_1 == 8 & X_2 == 2)) # & is "and"
```

. . .

-   How would you calculate it exactly using our `Pr()` function?

. . .

```{r}
Pr(x = 8, n = 10) * Pr(x = 2, n = 10 - 8)
```

. . .

-   Ancient (deterministic) calculations *describe* the Modern process by which the data in `frames` were simulated

## From [Aristotelian Logic](https://en.wikipedia.org/wiki/Boolean_algebra) to Probability

-   In R (and most other languages), `TRUE` maps to $1$ and `FALSE` maps to $0$ when doing arithmetic operations

```{r, AND}
c(TRUE & TRUE, TRUE & FALSE, FALSE & TRUE, FALSE & FALSE)
c(TRUE * TRUE, TRUE * FALSE, FALSE * TRUE, FALSE * FALSE)
```

. . .

::: incremental
-   Can generalize to numbers on the $[0,1]$ interval to calculate the probability that two (or more) propositions are both true. $\bigcap$ reads as "and". **General Multiplication Rule**: $\Pr\left(A\bigcap B\right)=\Pr\left(B\right)\times\Pr\left(A\mid B\right)=\Pr\left(A\right)\times\Pr\left(B\mid A\right)$
-   Iff $A$ and $B$ are independent, $\Pr\left(A \mid B\right) = \Pr\left(A\right)$ and $\Pr\left(B \mid A\right) = \Pr\left(B\right)$, so $\Pr\left(A\bigcap B\right) = \Pr\left(A\right) \times \Pr\left(B\right)$
:::

## Enumerating Bivariate Probabilities

```{r, joint_Pr}
#| code-line-numbers: 1|2-3|4-5|6-9
joint_Pr <- table(frames) / R # do this rather than vvv
joint_Pr <- matrix(0, nrow = length(Omega), ncol = length(Omega),
                   dimnames = list(Omega, Omega))
for (x_1 in Omega) {
  Pr_x_1 <- Pr(x_1, n = 10)
  for (x_2 in 0:(10 - x_1)) {
    joint_Pr[x_1 + 1, x_2 + 1] <- Pr_x_1 * Pr(x_2, n = 10 - x_1)
  } # R indexes starting from 1 (not 0), so have to +1 the indices
}
```

. . .

```{r}
sum(joint_Pr)
```

. . .

```{r}
#| eval: false
joint_Pr # do View(joint_Pr) to see it better than on the next slide
```

What is the probability of knocking down nine pins on the second roll of a frame of bowling, irrespective of what happens on the first roll? How can you utilize `joint_Pr` to calculate it?

##  {.smaller}

```{r}
#| echo: false
#| message: false
library(knitr)
library(kableExtra)
options("kableExtra.html.bsTable" = TRUE)
options(scipen = 5)
options(knitr.kable.NA = "")
tmp <- as.data.frame(joint_Pr)
for (i in 1:ncol(tmp)) 
  tmp[,i] <- cell_spec(round(tmp[,i], digits = 4), "html", 
                       bold = tmp[,i] == 0,
                       color = ifelse(tmp[,i] == 0, "red", "black"))
kable(tmp, format = "html", align = 'c', escape = FALSE,
      table.attr = 'data-quarto-disable-processing="true"') |>
  kable_styling("striped", full_width = FALSE)
```

## Probability of Non-Exclusive Events

-   How would you compute the probability that $X_1 = 8$ or $X_2 = 2$ in the same frame of bowling using `frames`?

. . .

```{r}
summarize(frames, # in R, | reads as "or" not "given" as in math
          prob  = mean(X_1 == 8 | X_2 == 2), 
          wrong = mean(X_1 == 8) + mean(X_2 == 2),
          right = mean(X_1 == 8) + mean(X_2 == 2) - 
                  mean(X_1 == 8 & X_2 == 2)) # fix double-counting
```

. . .

-   How would you calculate it exactly using `joint_Pr`?

. . .

```{r}
sum(joint_Pr["8", ]) + sum(joint_Pr[ , "2"]) - joint_Pr["8", "2"]
```

## Aristotelian Logic to Probability Again

```{r, OR}
c(TRUE | FALSE, FALSE | TRUE, FALSE | FALSE, TRUE | TRUE)
c(TRUE + FALSE, FALSE + TRUE, FALSE + FALSE, TRUE + TRUE - TRUE * TRUE)
```

::: incremental
-   Can generalize Aristotelian logic to numbers on the $[0,1]$ interval to calculate the probability that one of two propositions is true. $\bigcup$ is read as "or". **General Addition Rule**: $\Pr\left(A\bigcup B\right)=\Pr\left(A\right)+\Pr\left(B\right)-\Pr\left(A\bigcap B\right)$

-   If $\Pr\left(A\bigcap B\right) = 0$, $A$ and $B$ are said to be disjoint (or mutually exclusive)
:::

## A Function of a R.V. is a R.V.

::: incremental
-   What is the probability of getting a spare --- knocking down all $10$ pins over both rolls but not the first --- in bowling? How would you compute it using the simulated `frames`?
:::

. . .

```{r}
summarize(frames, spare_prob = mean(X_1 != 10 & X_1 + X_2 == 10))
```

. . .

-   How would you calculate it exactly using `joint_Pr`?

. . .

```{r}
spare_prob <- 0 # Have to +1 the indices in R
for (x_1 in 9:0) spare_prob <- spare_prob + joint_Pr[x_1 + 1, 10 - x_1 + 1]
spare_prob 
```

